---
title: "Saving Bioconductor Objects for Sharing"
author: "Michael Love"
date: today
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Saving Bioconductor Objects for Sharing}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Bioconductor objects are useful for their shared structure within the project, 
and that they enable rich metadata.

How then should one save a Bioconductor object so that a collaborator 
can load and use it? 
Or so that it persists reliably across time? 

The answer depends on several factors:

- Longevity: will the file need to be easily readable in 1 year, 
  or 5-10 years from now?
- Language interoperability: does the recipient use R, Python, or both?
- Object size: is the object small enough to serialize entirely, or does it
  contain large on-disk arrays?
- Is it easy enough to write out in common formats (BED) or plaintext?
- Reproducibility: should the saved form capture provenance and metadata
  alongside data?

This vignette walks through the main options, their trade-offs, and some
recommendations for common scenarios.

## R serialization

### saveRDS and readRDS

The simplest approach is R's built-in binary serialization. `saveRDS()` saves
a single object to an `.rds` file; `save()` bundles one or more named objects
into an `.RData` (or `.rda`) file.

```{r}
#| label: rds-save-load
#| message: false
library(SummarizedExperiment)

se <- SummarizedExperiment(
  assays = list(counts = matrix(1:12, nrow = 3)),
  colData = DataFrame(
    condition = c("A", "A", "B", "B"), 
    row.names=1:4
    ),
  rowData = DataFrame(
    gene = c("gene1","gene2","gene3"), 
    row.names=1:3
    )
)

# Single object
tmp_rds <- tempfile(fileext = ".rds")
saveRDS(se, file = tmp_rds)
se_from_rds <- readRDS(tmp_rds)
se_from_rds

# Multiple objects in one file
tmp_rda <- tempfile(fileext = ".RData")
save(se, file = tmp_rda)
load(tmp_rda)  # restores 'se' by name into the current environment
```

Prefer `saveRDS()` for most use cases. `save()` / `load()` silently overwrites
any object in the calling environment that shares a name, which is a common
source of confusion. The main remaining use case for `save()` is `.rda` data
files shipped inside R packages (under `data/`).

Advantages:

- Zero setup — works with any R object.
- Compression applied automatically (gzip by default).
- Round-trips perfectly: the loaded object is identical to the saved one.

Disadvantages:

- R-only: Python or other languages cannot read these files without a bridge
  library.
- Large objects are loaded entirely into RAM; there is no lazy / on-disk access.

When to use it: quick sharing between R users on the same project, saving
intermediate objects in a pipeline, anything under ~1 GB.

If your workflow downloads `.rds` files from a remote URL, consider using
[`BiocFileCache`](https://bioconductor.org/packages/BiocFileCache) to cache
them locally so they are only fetched once.

### Cross-release stability

There is no guarantee that an S4 object serialized today will work with older
versions of Bioconductor. The most common reasons are that the S4 class is not
defined at all in the earlier version, or that it exists but its definition has
changed — new slots added, slots renamed or removed, or infrastructure moved
between packages.

A concrete example: in a recent Bioconductor release, `Seqinfo` was moved out
of `GenomicRanges` into its own package. A `GRanges` serialized on a machine
with the newer setup could not be loaded on a machine with the older
`GenomicRanges` because the class definition for the embedded `Seqinfo` slot
was not found. The reverse was also true: older `GRanges` objects loaded into
a newer session sometimes required `updateObject()` to migrate the internal
representation:

```{r}
#| label: update-object
#| eval: false
# not evaluated — requires an object saved under an older Bioconductor release
gr <- readRDS("old_granges.rds")
gr <- updateObject(gr, verbose = TRUE)
```

The general lesson is: if you can save your data in a format that is not
tied to a particular version of Bioconductor — or better, not tied to R at
all — you should. For a `GRanges`, for instance, a simple TSV is often
enough:

```{r}
#| label: granges-tsv
#| message: false
library(GenomicRanges)

gr <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = c(100, 200, 300), width = 50)
)
names(gr) <- c("peak1", "peak2", "peak3")
gr$score  <- c(500, 800, 300)   # standard BED score column
gr$log2fc <- c(1.2, -0.5, 2.1) # extra metadata column

tmp_tsv <- tempfile(fileext = ".tsv")
write.table(as.data.frame(gr), tmp_tsv, sep = "\t", quote = FALSE)

gr_from_tsv <- makeGRangesFromDataFrame(
  read.table(tmp_tsv, header = TRUE, sep = "\t"),
  keep.extra.columns = TRUE
)
gr_from_tsv
```

This round-trip survives any Bioconductor version and is readable from Python
or the command line. The `alabaster` ecosystem (described below) applies the
same principle more systematically and with better support for complex objects,
using HDF5 and JSON as the underlying storage formats.

## HDF5-backed storage

### Saving with HDF5Array

For large assay matrices (e.g., single-cell count matrices with millions of
cells), it is impractical to hold the entire object in RAM. The
[`HDF5Array`](https://bioconductor.org/packages/HDF5Array) package provides
array classes backed by HDF5 files, enabling lazy loading and out-of-memory
computation.

```{r}
#| label: hdf5-save-load
#| message: false
library(HDF5Array)

tmp_hdf5 <- tempfile()
saveHDF5SummarizedExperiment(se, dir = tmp_hdf5, replace = TRUE)

# Assay data remains on disk until accessed
se_from_hdf5 <- loadHDF5SummarizedExperiment(tmp_hdf5)
se_from_hdf5
```

The saved directory contains an HDF5 file with the assay data and an `.rds`
file for the non-assay metadata.

Advantages:

- Assay data is stored on disk; only the chunks you access are read into RAM.
- HDF5 is a widely used binary format with readers in Python (`h5py`,
  `anndata`), Julia, C/C++, and more.
- Good for objects with tens of gigabytes of assay data.

Disadvantages:

- The output is a directory, not a single file, which complicates transfer
  (use `tar` or `zip` before sharing).
- The `.rds` envelope for metadata is still R-specific.
- Write performance can be slower than `saveRDS()` for small objects.

When to use it: large single-cell or spatial datasets where you want
on-disk access; workflows shared between R users who need memory efficiency.

### SummarizedExperiment and AnnData

For single-cell workflows that move between R and Python, the
[zellkonverter](https://bioconductor.org/packages/zellkonverter) package
converts directly between `SingleCellExperiment` and the Python `AnnData`
format (`.h5ad` files used by scanpy and related tools). Like `HDF5Array`,
`.h5ad` is an HDF5-based format.

```{r}
#| label: zellkonverter
#| eval: false
# not evaluated — zellkonverter installs a full Python environment via basilisk
# on first use, which takes too long in CI
library(zellkonverter)
library(SingleCellExperiment)

sce <- as(se, "SingleCellExperiment")

tmp_h5ad <- tempfile(fileext = ".h5ad")
writeH5AD(sce, file = tmp_h5ad)

sce_from_h5ad <- readH5AD(tmp_h5ad)
sce_from_h5ad
```

This is often the most convenient path when collaborators are working in
scanpy, as `.h5ad` is the native format on that side. The trade-off relative
to alabaster is that `.h5ad` is AnnData-specific rather than a general
Bioconductor serialization format.

## The alabaster ecosystem

### saveObject and readObject

The [`alabaster`](https://bioconductor.org/packages/alabaster.base) family of
packages is part of the broader [ArtifactDB](https://github.com/ArtifactDB)
project, which provides a multi-language system for storing and retrieving
analysis-ready Bioconductor objects. The core idea
is to save objects as directories of standard files (HDF5, JSON, CSV) whose
format is defined by explicit, versioned specifications — meaning the saved
form is readable without R, and can evolve over time without breaking
previously saved objects.

```{r}
#| label: alabaster-save-load
#| message: false
library(alabaster.base)
library(alabaster.se)

tmp_alabaster <- tempfile()
saveObject(se, path = tmp_alabaster)

se_from_alabaster <- readObject(tmp_alabaster)
se_from_alabaster
```

The `alabaster` umbrella package pulls in support for the most common
Bioconductor classes. Individual sub-packages cover specific classes:
`alabaster.se` for `SummarizedExperiment`, `alabaster.sce` for
`SingleCellExperiment`, and so on.

### Validation with takane

A key part of the ArtifactDB design is that saved directories can be
independently validated against the format specification. This is handled by
[takane](https://github.com/ArtifactDB/takane), a C++ library that maintains
separate, versioned specifications for 30+ Bioconductor object types. Calling
`takane::validate()` on a saved directory checks that all files conform to the
expected layout and types, which means a collaborator or downstream tool can
verify the integrity of a saved object without needing to load it into R.
This makes alabaster directories suitable for deposition in data repositories
where format conformance needs to be auditable.

The Python counterpart to `alabaster` is the
[dolomite](https://github.com/ArtifactDB/dolomite-base) family of packages,
which reads and writes the same on-disk format. An object saved with
`alabaster` in R can be read with `dolomite` in Python, and vice versa, with
no conversion step.

Advantages:

- Truly language-agnostic: Python readers exist via the `dolomite` family of
  packages, enabling seamless R ↔ Python interoperability.
- Built on open standards (HDF5, JSON); inspectable without R.
- Versioned format specifications with independent validation via takane.
- A good choice for archives or data portals.

Disadvantages:

- Newer ecosystem; not all Bioconductor classes have `alabaster` support yet.
- Requires installing the relevant `alabaster.*` sub-package for each class.
- Like HDF5Array, the output is a directory.

When to use it: archival storage, data portal submissions, cross-language
workflows, or any situation where you want the saved format to be readable
without R.

## BED format for ranges

### Writing BED files

When the object is a `GRanges` or similar ranges object and the goal is
interoperability with other tools (genome browsers, Python, command-line
utilities), exporting to BED format is often more useful than R-specific
serialization. Our `gr` has range names, a standard BED score column, and
an extra metadata column `log2fc`:

```{r}
#| label: show-gr
gr
```

Both `rtracklayer` and `plyranges` can write BED files:

```{r}
#| label: bed-write
#| message: false
library(rtracklayer)
library(plyranges)

tmp_bed <- tempfile(fileext = ".bed")
export(gr, tmp_bed)

tmp_bed2 <- tempfile(fileext = ".bed")
write_bed(gr, tmp_bed2)
```

When reading back, names and the standard `score` column are preserved, but
`log2fc` is silently dropped — BED has no mechanism to carry arbitrary
metadata columns:

```{r}
#| label: bed-read
gr_rtracklayer <- import(tmp_bed)
gr_rtracklayer

gr_plyranges <- read_bed(tmp_bed2)
gr_plyranges
```

### Saving metadata columns

To round-trip extra mcols, write them to a sidecar file. Here using plyranges
to read the BED back, then reattach from the sidecar:

```{r}
#| label: bed-mcols-sidecar
tmp_meta <- tempfile(fileext = ".tsv")
write.table(
  data.frame(name = names(gr), log2fc = gr$log2fc),
  tmp_meta, sep = "\t", quote = FALSE, row.names = FALSE
)

gr_restored <- read_bed(tmp_bed2)
meta <- read.table(tmp_meta, header = TRUE, sep = "\t")
gr_restored$log2fc <- meta$log2fc
gr_restored
```

### Saving Seqinfo separately

BED files do not store chromosome lengths or genome build information, so
`Seqinfo` is silently dropped on export. To preserve it, write it out
alongside the BED file and restore it on load:

```{r}
#| label: seqinfo-save-restore
tmp_seqinfo <- tempfile(fileext = ".csv")
write.csv(as.data.frame(seqinfo(gr)), tmp_seqinfo)

df <- read.csv(tmp_seqinfo, row.names = 1)
si <- Seqinfo(
  seqnames   = rownames(df),
  seqlengths = as.integer(df$seqlengths),
  isCircular = as.logical(df$isCircular),
  genome     = as.character(df$genome)
)
si
```

## Propagating object metadata

Object-level metadata stored in `metadata(object)` — things like processing
parameters, provenance notes, or experiment descriptors — is lost in any
format that only encodes the ranges or assay data. This applies whether you
are writing a BED file, an HDF5 matrix, or any other non-R format. Write it
to a JSON sidecar file so it travels with the data:

```{r}
#| label: metadata-json
#| message: false
library(jsonlite)

metadata(se) <- list(genome = "hg38", pipeline = "v2.1", n_samples = 4L)

tmp_json <- tempfile(fileext = ".json")
writeLines(toJSON(metadata(se), pretty = TRUE, auto_unbox = TRUE), tmp_json)

metadata(se_from_rds) <- fromJSON(tmp_json)
metadata(se_from_rds)
```

`toJSON` handles simple R types (lists, vectors, data frames) well, but
complex objects (S4 instances, environments) need to be simplified or omitted
before serializing.

## Summary and recommendations

| Scenario | Recommended approach |
|---|---|
| Quick sharing between R users, same Bioc release | `saveRDS()` |
| Loading an object from an older Bioc release | `updateObject()` after `readRDS()` |
| Large assay matrices, R-only | `saveHDF5SummarizedExperiment()` |
| SingleCellExperiment ↔ Python AnnData | `zellkonverter::writeH5AD()` / `readH5AD()` |
| Cross-language (R + Python), general | `alabaster::saveObject()` |
| Long-term archive / data portal | `alabaster::saveObject()` |

In most new projects we recommend defaulting to `saveRDS()` for convenience
and upgrading to `alabaster` when cross-language access or archival stability
becomes a priority. Be aware that any R-serialized Bioconductor object may
require `updateObject()` when loaded under a different Bioconductor release.

## Session info

```{r}
#| label: session-info
sessionInfo()
```
