---
title: "Saving Bioconductor Objects for Sharing"
author: "Mike Love"
date: today
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Saving Bioconductor Objects for Sharing}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Introduction

A common question in Bioinformatics workflows is: how should I save a
Bioconductor object so that a collaborator can load and use it, or so that it
persists reliably across time? The answer depends on several factors:

- Longevity: will the file need to be readable in 5 or 10 years?
- Language interoperability: does the recipient use R, Python, or both?
- Object size: is the object small enough to serialize entirely, or does it
  contain large on-disk arrays?
- Reproducibility: should the saved form capture provenance and metadata
  alongside data?

This vignette walks through the main options, their trade-offs, and our
recommendations for common scenarios.

## R serialization

The simplest approach is R's built-in binary serialization. `saveRDS()` saves
a single object to an `.rds` file; `save()` bundles one or more named objects
into an `.RData` (or `.rda`) file.

```{r}
#| eval: false
se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = matrix(1:12, nrow = 3)),
  colData = S4Vectors::DataFrame(condition = c("A", "A", "B", "B"))
)

# Single object
saveRDS(se, file = "my_se.rds")
se2 <- readRDS("my_se.rds")

# Multiple objects in one file
save(se, file = "my_se.RData")
load("my_se.RData")  # 'se' appears in the workspace
```

Prefer `saveRDS()` for most use cases. `save()` / `load()` silently overwrites
any object in the calling environment that shares a name, which is a common
source of confusion. The main remaining use case for `save()` is `.rda` data
files shipped inside R packages (under `data/`).

Advantages:

- Zero setup — works with any R object.
- Compression applied automatically (gzip by default).
- Round-trips perfectly: the loaded object is identical to the saved one.

Disadvantages:

- R-only: Python or other languages cannot read these files without a bridge
  library.
- Large objects are loaded entirely into RAM; there is no lazy / on-disk access.

When to use it: quick sharing between R users on the same project, saving
intermediate objects in a pipeline, anything under ~1 GB.

If your workflow downloads `.rds` files from a remote URL, consider using
[`BiocFileCache`](https://bioconductor.org/packages/BiocFileCache) to cache
them locally so they are only fetched once.

### Cross-release stability

There is no guarantee that an S4 object serialized today will work with older
versions of Bioconductor. The most common reasons are that the S4 class is not
defined at all in the earlier version, or that it exists but its definition has
changed — new slots added, slots renamed or removed, or infrastructure moved
between packages.

A concrete example: in a recent Bioconductor release, `Seqinfo` was moved out
of `GenomicRanges` into its own package. A `GRanges` serialized on a machine
with the newer setup could not be loaded on a machine with the older
`GenomicRanges` because the class definition for the embedded `Seqinfo` slot
was not found. The reverse was also true: older `GRanges` objects loaded into
a newer session sometimes required `updateObject()` to migrate the internal
representation:

```{r}
#| eval: false
gr <- readRDS("old_granges.rds")
gr <- updateObject(gr, verbose = TRUE)
```

The general lesson is: if you can save your data in a format that is not
tied to a particular version of Bioconductor — or better, not tied to R at
all — you should. For a `GRanges`, for instance, a simple TSV is often
enough:

```{r}
#| eval: false
# Save
write.table(as.data.frame(gr), "ranges.tsv", sep = "\t", quote = FALSE)

# Restore
gr2 <- makeGRangesFromDataFrame(
  read.table("ranges.tsv", header = TRUE, sep = "\t"),
  keep.extra.columns = TRUE
)
```

This round-trip survives any Bioconductor version and is readable from Python
or the command line. The `alabaster` ecosystem (described below) applies the
same principle more systematically and with better support for complex objects,
using HDF5 and JSON as the underlying storage formats.

## HDF5-backed storage

For large assay matrices (e.g., single-cell count matrices with millions of
cells), it is impractical to hold the entire object in RAM. The
[`HDF5Array`](https://bioconductor.org/packages/HDF5Array) package provides
array classes backed by HDF5 files, enabling lazy loading and out-of-memory
computation.

```{r}
#| eval: false
library(HDF5Array)
library(SummarizedExperiment)

# Save: writes a .h5 file and a small .rds "envelope"
saveHDF5SummarizedExperiment(se, dir = "my_se_hdf5/", replace = TRUE)

# Load: the assay data remains on disk until accessed
se_h5 <- loadHDF5SummarizedExperiment("my_se_hdf5/")
```

The saved directory contains an HDF5 file with the assay data and an `.rds`
file for the non-assay metadata.

Advantages:

- Assay data is stored on disk; only the chunks you access are read into RAM.
- HDF5 is a widely used binary format with readers in Python (`h5py`,
  `anndata`), Julia, C/C++, and more.
- Good for objects with tens of gigabytes of assay data.

Disadvantages:

- The output is a directory, not a single file, which complicates transfer
  (use `tar` or `zip` before sharing).
- The `.rds` envelope for metadata is still R-specific.
- Write performance can be slower than `saveRDS()` for small objects.

When to use it: large single-cell or spatial datasets where you want
on-disk access; workflows shared between R users who need memory efficiency.

## The alabaster ecosystem

The [`alabaster`](https://bioconductor.org/packages/alabaster.base) family of
packages provides a language-agnostic, file-based serialization format.
Objects are saved as a directory of standard files (HDF5, JSON, CSV) with a
schema that can be read from R, Python, or any other language.

```{r}
#| eval: false
library(alabaster.base)

# Save to a directory
saveObject(se, path = "my_se_alabaster/")

# Load back
se3 <- readObject("my_se_alabaster/")
```

The `alabaster` umbrella package pulls in support for the most common
Bioconductor classes. Individual sub-packages cover specific classes:
`alabaster.se` for `SummarizedExperiment`, `alabaster.sce` for
`SingleCellExperiment`, and so on.

Advantages:

- Truly language-agnostic: Python readers exist via the `dolomite` family of
  packages, enabling seamless R ↔ Python interoperability.
- Built on open standards (HDF5, JSON); inspectable without R.
- Forward-designed for long-term reproducibility and schema versioning.
- A good choice for archives or data portals.

Disadvantages:

- Newer ecosystem; not all Bioconductor classes have `alabaster` support yet.
- Requires installing the relevant `alabaster.*` sub-package for each class.
- Like HDF5Array, the output is a directory.

When to use it: archival storage, data portal submissions, cross-language
workflows, or any situation where you want the saved format to be readable
without R.

### SummarizedExperiment and AnnData

For single-cell workflows that move between R and Python, a common alternative
to alabaster is
[zellkonverter](https://bioconductor.org/packages/zellkonverter), which
converts directly between `SingleCellExperiment` and the Python `AnnData`
format (`.h5ad` files used by scanpy and related tools).

```{r}
#| eval: false
library(zellkonverter)

# SingleCellExperiment -> .h5ad (readable by scanpy in Python)
writeH5AD(sce, file = "my_sce.h5ad")

# .h5ad -> SingleCellExperiment
sce2 <- readH5AD("my_sce.h5ad")
```

This is often the most convenient path when collaborators are working in
scanpy, as `.h5ad` is the native format on that side. The trade-off relative
to alabaster is that `.h5ad` is AnnData-specific rather than a general
Bioconductor serialization format.

## Exporting genomic ranges to BED

When the object is a `GRanges` or similar ranges object and the goal is
interoperability with other tools (genome browsers, Python, command-line
utilities), exporting to BED format is often more useful than R-specific
serialization. Two options:

```{r}
#| eval: false
# rtracklayer — general-purpose genomics I/O
library(rtracklayer)
export(gr, "ranges.bed")

# plyranges — tidier interface, writes BED3/BED6 depending on the object
library(plyranges)
write_bed(gr, "ranges.bed")
```

### Saving Seqinfo separately

BED files do not store chromosome lengths or genome build information, so
`Seqinfo` is silently dropped on export. To preserve it, write it out
alongside the BED file and restore it on load:

```{r}
#| eval: false
# Save
write.csv(as.data.frame(seqinfo(gr)), "seqinfo.csv")

# Restore
si <- as(read.csv("seqinfo.csv", row.names = 1), "Seqinfo")
seqinfo(gr2) <- si
```

## Saving metadata to JSON

Object-level metadata stored in `metadata(object)` — things like processing
parameters, provenance notes, or experiment descriptors — is lost in any
format that only encodes the ranges or assay data. This applies whether you
are writing a BED file, an HDF5 matrix, or any other non-R format. Write it
to a JSON sidecar file so it travels with the data:

```{r}
#| eval: false
library(jsonlite)

# Write
writeLines(toJSON(metadata(se), pretty = TRUE, auto_unbox = TRUE),
           "metadata.json")

# Read back
metadata(se2) <- fromJSON("metadata.json")
```

`toJSON` handles simple R types (lists, vectors, data frames) well, but
complex objects (S4 instances, environments) need to be simplified or omitted
before serializing.

## Summary and recommendations

| Scenario | Recommended approach |
|---|---|
| Quick sharing between R users, same Bioc release | `saveRDS()` |
| Shipping small data in a package | `save()` → `data/*.rda` |
| Large assay matrices, R-only | `saveHDF5SummarizedExperiment()` |
| Cross-language (R + Python) | `alabaster::saveObject()` |
| Long-term archive / data portal | `alabaster::saveObject()` |
| Sharing genomic ranges with non-R tools | `rtracklayer::export()` or `plyranges::write_bed()` |
| Loading an object from an older Bioc release | `updateObject()` after `readRDS()` |

In most new projects we recommend defaulting to `saveRDS()` for convenience
and upgrading to `alabaster` when cross-language access or archival stability
becomes a priority. Be aware that any R-serialized Bioconductor object may
require `updateObject()` when loaded under a different Bioconductor release.

## Session info

```{r}
sessionInfo()
```
